---
title: "t-test"
author: "spike"
date: "2025-04-07"
output: ioslides_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  # echo = FALSE, 
  warning = FALSE, 
  error = FALSE, 
  message = FALSE,
  fig.align = "center",
  out.width = "80%"
  )
```

```{r, echo=FALSE}
library(tidyverse)
theme_set(jtools::theme_apa())
```

## Recap Lecture: the t-test

<!--
comment template
-->

t-test as a case of proof by contradiction

| phase | proof by contradiction | t-test |
|-------|-----------------------|--------|
| assume | all sheep are white | $H_0$: $\mu = 0$ |
| test | find a black sheep | $p(\bar{x} | \mu = 0)$ |
| reject | contradiction | $p = 0.0001$, unlikely! |
| conclude | some sheep are black | $h_a$: $\mu \ne 0$ |

## Step1: all sheep are white

$H_0$: the population mean is 0. So, in perfect world:

```{r, echo=FALSE}
config <- tibble(
  x_len = 1000,
  N = 5,
)

density_df <-
  with(
    config,
    {
      population <- tibble(
        x = seq(-3, 3, length.out = x_len),
        y = dnorm(x),
        name = "population"
      )
      
      t_dist <- tibble(
        x = seq(-3, 3, length.out = x_len),
        y = dt(x, df = N - 1),
        name = "sample_mean"
      )
      
      population %>%
        bind_rows(t_dist)
    }
  )

density_df %>%
ggplot(aes(x = x, y = y)) +
      geom_line() +
      facet_grid(~ name, scales = "free")
```

## Step1: all sheep are white

If we draw a sample of 5:

```{r, echo=FALSE}
set.seed(123)
sample_df <- 
  with(
    config,
    {
      tibble(
        x = rnorm(N, mean = 0, sd = 1),
        name = "population"
      )
    }
  )

ggplot() +
  geom_line(
    data = density_df,
    aes(x = x, y = y),
    alpha = 0.5
  ) +
  facet_grid(~ name, scales = "free") +
  geom_point(
    data = sample_df,
    aes(x = x, y = 0),
    size = 1,
    color = "red"
  ) +
  geom_vline(
    data = sample_df %>%
      summarise(
        mean = mean(x),
        name = "sample_mean"
      ),
    aes(xintercept = mean),
    color = "blue",
    linetype = "dashed"
  ) 
```

## Step2: find a black sheep

What if we find a black sheep?

- Think about how is it plausible that we find this if our assumption is true?

```{r, echo=FALSE}
set.seed(123)
sample_df <- 
  with(
    config,
    {
      tibble(
        x = rnorm(N, mean = 2, sd = 1),
        name = "population"
      )
    }
  )

ggplot() +
  geom_line(
    data = density_df,
    aes(x = x, y = y),
    alpha = 0.5
  ) +
  facet_grid(~ name, scales = "free") +
  geom_point(
    data = sample_df,
    aes(x = x, y = 0),
    size = 1,
    color = "red"
  ) +
  geom_vline(
    data = sample_df %>%
      summarise(
        mean = mean(x),
        name = "sample_mean"
      ),
    aes(xintercept = mean),
    color = "blue",
    linetype = "dashed"
  ) 
```

## Step3: reject the null hypothesis

I cannot believe that they are randomly drawn from a population with mean 0.

```{r, echo=FALSE}
set.seed(123)
sample_df <- 
  with(
    config,
    {
      tibble(
        x = rnorm(N, mean = 2, sd = 1),
        name = "population"
      )
    }
  )

ggplot() +
  geom_line(
    data = density_df,
    aes(x = x, y = y),
    alpha = 0.5
  ) +
  facet_grid(~ name, scales = "free") +
  geom_point(
    data = sample_df,
    aes(x = x, y = 0),
    size = 1,
    color = "red"
  ) +
  geom_vline(
    data = sample_df %>%
      summarise(
        mean = mean(x),
        name = "sample_mean"
      ),
    aes(xintercept = mean),
    color = "blue",
    linetype = "dashed"
  ) +
  geom_ribbon(
    data = density_df %>%
      filter(name == "sample_mean") %>%
      filter(x < qnorm(0.95)),
    aes(x = x, ymin = 0, ymax = y),
    fill = "blue",
    alpha = 0.2
  ) +
  geom_text(
    data = tibble(content = "95% lower tail", name = "sample_mean"),
    aes(x = 0, y = 0.1, label = content),
    size = 5,
    color = "black",
  )
```

## Three types of t-tests

| phase | one-sample | paired | independent |
|-------|------------|--------|-------------|
| want | $H_1$: $\mu \ne 0$ | $H_1$: $\mu_{\text{diff}} \ne 0$ | $H_1$: $\mu_1 \ne \mu_2$ |
| assume | $H_0$: $\mu = 0$ | $H_0$: $\mu_{\text{diff}} = 0$ | $H_0$: $\mu_1 = \mu_2$ |

- paired $\Leftrightarrow$  one-sample: paired is a special case of one-sample.

## Recap Lecture: the effect size

A fancy way to say: if we find a black sheep, how black is it?

- So, it is significant, how big is the difference?
- Example: if a group of people have IQ of 101, which is significant, but not very interesting.

## Recap Lecture: the effect size

Find the similarity:

$$
\text{Cohen’s } d = \frac{\bar{x}_1 - \bar{x}_2}{s_{\text{pooled}}}
$$

$$
\text{Z-score} = \frac{\bar{x} - \mu}{s_{\text{sample}}}
$$

dividing by standard deviation is the common trick to standardize the magnitude.
