---
title: "Descriptive Statistics and the Normal Distribution"
author: "spike"
date: "2025-03-10"
output: ioslides_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  # echo = FALSE, 
  warning = FALSE, 
  error = FALSE, 
  message = FALSE
  )
```

```{r, echo=FALSE}
library(tidyverse)
library(gganimate)
theme_set(jtools::theme_apa())
```

## Central Tendency

| index | 1 | 2 | 3 | 4 | 5 |
|---|---|---|---|---|---|
|X | 2 | 3 | 4 | 4 | 5 |

- Mean: $\frac{\sum X}{N} = \frac{2 + 3 + 4 + 4 + 5}{5} = 3.6$
- Median: third value = 4
- Mode: Most frequent value = 4

## Central Tendency

Why we need central tendency? Consider a group of students with the following marks:

```{r, echo=FALSE}
marks_params <- tibble(
  N = 18,
  mu = 0.7,
  phi = 5,
  alpha = mu * phi,
  beta = (1 - mu) * phi
)

set.seed(123)
marks <- with(
  marks_params,
  {
    marks <- tibble(
      mark = round(rbeta(N, alpha, beta) * 100, 1),
      selected = FALSE
    ) %>%
      bind_rows(
        tibble(
          mark = mu * 100,
          selected = TRUE
        )
      )
    marks
  }
)

marks %>%
  ggplot(aes(x = mark)) +
  geom_point(y = 0.5, alpha = 0.5) +
  xlim(0, 100)
```

## Central Tendency

```{r, echo=FALSE}
marks %>%
  ggplot(aes(x = mark)) +
  geom_point(y = 0.5, alpha = 0.5) +
  geom_vline(aes(xintercept = mean(mark)), color = "red") +
  geom_text(
    aes(x = mean(mark), y = 0.8, label = "Mean"),
    color = "red"
  ) +
  geom_vline(aes(xintercept = median(mark)), color = "blue") +
  geom_text(
    aes(x = median(mark), y = 0.7, label = "Median"),
    color = "blue"
  ) +
  ylim(0, 1) +
  theme(
    axis.ticks.y = element_blank(),
    axis.text.y = element_blank(),
    axis.title.y = element_blank()
  ) +
  xlim(0, 100)
```

## Central Tendency - Manipulate One Score

```{r, echo=FALSE}
marks %>%
  ggplot(aes(x = mark)) +
  geom_point(y = 0.5, alpha = 0.5, aes(color = selected)) +
  scale_color_manual(values = c("TRUE" = "red", "FALSE" = "black")) +
  geom_vline(aes(xintercept = mean(mark)), color = "red") +
  geom_text(
    aes(x = mean(mark), y = 0.8, label = "Mean"),
    color = "red"
  ) +
  geom_vline(aes(xintercept = median(mark)), color = "blue") +
  geom_text(
    aes(x = median(mark), y = 0.7, label = "Median"),
    color = "blue"
  ) +
  ylim(0, 1) +
  theme(
    axis.ticks.y = element_blank(),
    axis.text.y = element_blank(),
    axis.title.y = element_blank(),
    legend.position = "none"
  ) +
  xlim(0, 100)
```

## Central Tendency - Manipulate One Score

```{r, echo=FALSE}
marks %>%
  expand_grid(
    tibble(
      manipulated = seq(1, 99)
    )
  ) %>%
  mutate(
    mark = if_else(selected, manipulated, mark),
  ) %>%
  group_by(manipulated) %>%
  mutate(
    mark_mean = mean(mark),
    mark_median = median(mark)
  ) %>%
  ggplot(aes(x = mark)) +
  geom_point(y = 0.5, alpha = 0.5, aes(color = selected)) +
  scale_color_manual(values = c("TRUE" = "red", "FALSE" = "black")) +
  geom_vline(aes(xintercept = mark_mean), color = "red") +
  geom_text(
    aes(x = mark_mean, y = 0.8, label = "Mean"),
    color = "red"
  ) +
  geom_vline(aes(xintercept = mark_median), color = "blue") +
  geom_text(
    aes(x = mark_median, y = 0.7, label = "Median"),
    color = "blue"
  ) +
  ylim(0, 1) +
  theme(
    axis.ticks.y = element_blank(),
    axis.text.y = element_blank(),
    axis.title.y = element_blank(),
    legend.position = "none"
  ) +
  xlim(0, 100) +
  transition_states(manipulated)
```

## Dispersion Tendency

| index | 1 | 2 | 3 | 4 | 5 |
|---|---|---|---|---|---|
|X | 2 | 3 | 4 | 4 | 5 |

- Variance: $\sigma^2_X = \frac{\sum (X - \mu_X)^2}{N}$
$= \frac{(2-3.6)^2 + (3-3.6)^2 + (4-3.6)^2 + (4-3.6)^2 + (5-3.6)^2}{5} = 1.04$
- Standard Deviation: $\sigma_X = \sqrt{\sigma^2_X} = \sqrt{1.04} = 1.02$

## Dispersion Tendency - Intuition

$\sigma^2_X = \frac{\sum (X - \mu_X)^2}{N}$

- $X_i - \mu_X$: Distance
- $(X_i - \mu_X)^2$: Positive Distance Measure
- $\frac{\sum (X - \mu_X)^2}{N}$: Average Positive Distance Measure

## Dispersion Tendency - Intuition

| index | 1 | 2 | 3 | 4 |
|---|---|---|---|---|
|X | 3 | 5 | 8 | 9 |

```{r, echo=FALSE}
dispersion_demo_df <- tribble(
  ~x, ~y,
  3, 4,
  5, 5,
  8, 6,
  9, 7
)

dispersion_demo_config <- summarise(
  .data = dispersion_demo_df,
  mean_y = mean(y),
)

dispersion_demo_df %>%
  ggplot(aes(x = x, y = mean(y))) +
  geom_point() +
  geom_vline(aes(xintercept = mean(x)), color = "red", linetype = "dashed") +
  geom_text(
    aes(x = mean(x), y = 0.35, label = "Mean"),
    color = "red"
  ) +
  ylim(dispersion_demo_config$mean_y - 5, dispersion_demo_config$mean_y + 5) +
  xlim(0, 10) +
  theme(
    axis.ticks.y = element_blank(),
    axis.text.y = element_blank(),
    axis.title.y = element_blank()
  ) +
  coord_fixed()
```

## Dispersion Tendency - Intuition

```{r, echo=FALSE}
dispersion_demo_df %>%
  ggplot(aes(x = x, y = y)) +
  geom_point() +
  geom_vline(aes(xintercept = mean(x)), color = "red", linetype = "dashed") +
  ylim(dispersion_demo_config$mean_y - 5, dispersion_demo_config$mean_y + 5) +
  xlim(0, 10) +
  theme(
    axis.ticks.y = element_blank(),
    axis.text.y = element_blank(),
    axis.title.y = element_blank()
  ) +
  coord_fixed()
```

## Dispersion Tendency - Intuition

$$
X - \mu_X
$$

```{r, echo=FALSE}
dispersion_demo_df %>%
  ggplot(aes(x = x, y = y)) +
  geom_point() +
  geom_vline(aes(xintercept = mean(x)), color = "red", linetype = "dashed") +
  geom_segment(aes(x = x, xend = mean(x), y = y, yend = y), color = "blue") +
  ylim(dispersion_demo_config$mean_y - 5, dispersion_demo_config$mean_y + 5) +
  xlim(0, 10) +
  theme(
    axis.ticks.y = element_blank(),
    axis.text.y = element_blank(),
    axis.title.y = element_blank()
  ) +
  coord_fixed()
```

## Dispersion Tendency - Intuition

$$
(X - \mu_X)^2
$$

```{r, echo=FALSE}
dispersion_sq_df <- tribble(
  ~x, ~y, ~xmin, ~xmax, ~ymin, ~ymax,
  3, 4, 3, 3 + 3.25, 4 - 3.25, 4,
  5, 5, 5, 5 + 1.25, 5, 5 + 1.25,
  8, 6, 8 - 1.75, 8, 6 - 1.75, 6,
  9, 7, 9 - 2.75, 9, 7, 7 + 2.75
)

dispersion_demo_df %>%
  mutate(
    side_len = abs(x - mean(x))
  ) %>%
  ggplot(aes(x = x, y = y)) +
  geom_point() +
  geom_vline(aes(xintercept = mean(x)), color = "red", linetype = "dashed") +
  geom_segment(aes(x = x, xend = mean(x), y = y, yend = y), color = "blue") +
  geom_rect(aes(
    xmin = xmin,
    xmax = xmax,
    ymin = ymin,
    ymax = ymax
  ), fill = "blue", alpha = 0.3, data = dispersion_sq_df) +
  ylim(dispersion_demo_config$mean_y - 5, dispersion_demo_config$mean_y + 5) +
  xlim(0, 10) +
  theme(
    axis.ticks.y = element_blank(),
    axis.text.y = element_blank(),
    axis.title.y = element_blank()
  ) +
  coord_fixed()
```

## Dispersion Tendency - Intuition

```{r, echo=FALSE}
dispersion_demo_df %>%
  mutate(
    side_len = abs(x - mean(x))
  ) %>%
  ggplot(aes(x = x, y = y)) +
  geom_point() +
  geom_vline(aes(xintercept = mean(x)), color = "red", linetype = "dashed") +
  geom_segment(aes(x = x, xend = mean(x), y = y, yend = y), color = "blue") +
  geom_rect(aes(
    xmin = xmin,
    xmax = xmax,
    ymin = ymin,
    ymax = ymax
  ), fill = "darkblue", data = tribble(
    ~x, ~y, ~xmin, ~xmax, ~ymin, ~ymax,
    0, 0, 0, 2.38, 8, 8 + 2.38
  )) +
  geom_text(
    aes(x = 1.19, y = 7.5, label = "Variance"),
    color = "darkblue"
  ) +
  geom_rect(aes(
    xmin = xmin,
    xmax = xmax,
    ymin = ymin,
    ymax = ymax
  ), fill = "blue", alpha = 0.3, data = dispersion_sq_df) +
  ylim(dispersion_demo_config$mean_y - 5, dispersion_demo_config$mean_y + 5) +
  xlim(0, 10) +
  theme(
    axis.ticks.y = element_blank(),
    axis.text.y = element_blank(),
    axis.title.y = element_blank()
  ) +
  coord_fixed()
```

## Practice

Now, let's finish Q1 - Q5. 

Questions 1 to 5 are intended to be done by hand. We will check the results in R subsequently.

## pnorm

```{r, echo=FALSE}
visualize_pnorm <- function(q_value) {
  tibble(
    x = seq(-4, 4, 0.01),
    d = dnorm(x)
  ) %>%
    ggplot(aes(x = x)) +
    geom_line(aes(y = d), color = "blue") +
    geom_ribbon(
      aes(ymin = 0, ymax = d),
      data = . %>% filter(x <= q_value),
      alpha = 0.3,
      fill = "red",
    ) +
    geom_vline(xintercept = q_value) +
    geom_text(
      x = q_value,
      hjust = 1,
      vjust = 1,
      y = 0.1,
      label = paste0("pnorm(", q_value, ") = ", round(pnorm(q_value), 2)),
      color = "red",
    ) +
    labs(
      x = "z-score",
      y = "",
      title = paste0("Given z-score (black line): ", q_value, ", what is the probability (red)?"),
    ) +
    theme(
      axis.ticks.y = element_blank(),
      axis.text.y = element_blank(),
      axis.title.y = element_blank()
    )
}

visualize_qnorm <- function(p_value) {
  tibble(
    x = seq(-4, 4, 0.01),
    d = dnorm(x)
  ) %>%
    ggplot(aes(x = x)) +
    geom_line(aes(y = d), color = "blue") +
    geom_vline(xintercept = qnorm(p_value), color = "red") +
    geom_ribbon(
      aes(ymin = 0, ymax = d),
      data = . %>% filter(x <= qnorm(p_value)),
      alpha = 0.3,
    ) +
    geom_text(
      x = qnorm(p_value),
      y = 0.1,
      hjust = 1,
      vjust = 1,
      label = paste0("qnorm(", p_value, ") = ", round(qnorm(p_value), 2)),
      color = "red",
    ) +
    labs(
      x = "z-score",
      y = "",
      title = paste0("Given probability (gray area): ", p_value, ", what is the z-score (red)?"),
    ) +
    theme(
      axis.ticks.y = element_blank(),
      axis.text.y = element_blank(),
      axis.title.y = element_blank()
    )
}
```

```{r, echo=FALSE}
pnorm(1.28)
visualize_pnorm(1.28)
```

## qnorm

```{r, echo=FALSE}
qnorm(0.95)
visualize_qnorm(0.95)
```